---
  title: "Gene signature design"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(teal)
library(teal.modules.hermes)
library(hermes)
library(checkmate)
```

## Objectives

- when multiple genes are selected, then prompt the user how to combine them into a signature

## Hermes Interface

It seems most natural to do the gene signature calculation within the `hermes` functions. Otherwise we would need to hack it by adding another pseudo-gene to all the assays in the experiment which contains the signature values.

### Draw function

Just as an example here a simplified boxplot function.

1. We add the `summary_fun` argument which takes a matrix function, typically starting with the prefix `col`, e.g. `colMeans()`. This is analogous to what we have in `top_genes()` where we take a matrix function operating on `row` space.
1. Now we take potentially multiple y variables, therefore rename the argument to plural and adapt the assertion accordingly.
1. We also need to take care of the label for the signature.

For the last one we need a little helper.

```{r}
shorten_list <- function(x) {
  n <- length(x)
  if (n > 3) {
    x <- c(x[1:2], "...", x[n])
  }
  paste(x, collapse = ", ")
}
shorten_list(letters[1:4])
shorten_list(letters)

parenthesize <- function(x) {
  if (is_blank(x)) {
    ""
  } else {
    paste0("(", x, ")")
  }
}
```

This will make our life easier below.

```{r}
draw_signature <- function(object,
                           assay_name,
                           x_var,
                           y_vars,
                           summary_fun = colMeans,
                           fun_name = deparse(substitute(summary_fun))) {
  assert_class(object, "AnyHermesData")
  assert_string(assay_name)
  assert_string(x_var)
  assert_character(y_vars, min.len = 1L, any.missing = FALSE, unique = TRUE)
  assay_matrix <- assay(object, assay_name)

  col_data <- colData(object)
  assert_names(names(col_data), must.include = x_var)
  assert_names(rownames(assay_matrix), must.include = y_vars)

  y_vals <- if (length(y_vars) > 1L) {
    assert_function(summary_fun)
    summary_fun(assay_matrix[y_vars, ])
  } else {
    assay_matrix[y_vars, ]
  }

  y_label <- if (length(y_vars) > 1L) {
    assert_string(fun_name, min.chars = 1L)
    paste0(fun_name, parenthesize(shorten_list(y_vars)))
  } else {
    y_vars
  }

  df <- data.frame(x = col_data[, x_var], y = y_vals)

  p <- ggplot(df, aes(x = .data$x, y = .data$y)) +
    geom_boxplot(outlier.shape = 19) +
    stat_boxplot(geom = "errorbar") +
    labs(x = x_var, y = y_label)
  geom_point_args <- list()
  p <- p + do.call(geom_point, geom_point_args)
  p
}
```

Let's try it out.

With one gene:

```{r}
hd <- HermesData(summarized_experiment)
draw_signature(
  hd,
  assay_name = "counts",
  y_var = genes(hd)[1],
  x_var = "COUNTRY"
)
```

With multiple (here 10) genes:

```{r}
draw_signature(
  hd,
  assay_name = "counts",
  y_var = genes(hd)[1:10],
  x_var = "COUNTRY"
)
```


## Teal Module Prototype 1

### Module function 

Here we allow the user to modify the list of gene signature functions.

```{r}
tm_g_signature <- function(label,
                           mae_name,
                           summary_funs = list(
                             Mean = colMeans,
                             Median = colMedians,
                             Max = colMaxs
                           ),
                           pre_output = NULL,
                           post_output = NULL) {
  assert_string(label)
  assert_string(mae_name)
  assert_list(
    summary_funs,
    types = "function",
    min.len = 1L,
    unique = TRUE,
    any.missing = FALSE,
    names = "unique"
  )
  assert_tag(pre_output, null.ok = TRUE)
  assert_tag(post_output, null.ok = TRUE)

  module(
    label = label,
    server = srv_g_signature,
    server_args = list(
      mae_name = mae_name,
      summary_funs = summary_funs
    ),
    ui = ui_g_signature,
    ui_args = list(
      mae_name = mae_name,
      summary_funs = summary_funs,
      pre_output = pre_output,
      post_output = post_output
    ),
    datanames = "all"
  )
}
```

### UI function

Here we take the `summary_funs` list and give that into the option menu.

```{r}
ui_g_signature <- function(id,
                           datasets,
                           mae_name,
                           summary_funs,
                           pre_output,
                           post_output) {
  ns <- NS(id)
  mae <- datasets$get_data(mae_name, filtered = FALSE)
  experiment_name_choices <- names(mae)

  tagList(
    include_css_files(pattern = "*"),
    teal.widgets::standard_layout(
      encoding = div(
        selectInput(ns("experiment_name"), "Select experiment", experiment_name_choices),
        selectInput(ns("assay_name"), "Select assay", choices = ""),
        div(
          class = "row",
          div(
            class = "col-sm-8",
            tags$label(
              class = "control-label",
              "Select gene(s) of interest"
            )
          ),
          div(
            class = "col-sm-4",
            actionButton(
              ns("open_text_input"),
              span(icon("font fa-border")),
              title = "Enter list of genes",
              class = "pull-right list-genes"
            ),
            div(
              class = "pull-right",
              title = "Lock gene selection",
              shinyWidgets::prettyToggle(
                ns("lock"),
                value = FALSE,
                label_on = NULL,
                label_off = NULL,
                status_on = "default",
                status_off = "default",
                outline = FALSE,
                plain = TRUE,
                icon_on = icon("lock fa-border"),
                icon_off = icon("unlock-alt fa-border"),
                animation = "pulse"
              )
            )
          )
        ),
        div(
          class = "custom-select-input",
          teal.widgets::optionalSelectInput(
            ns("y_vars"),
            label = NULL,
            choices = "",
            multiple = TRUE
          )
        ),
        conditionalPanel(
          condition = "input.y_vars.length > 1", # JavaScript expression
          ns = ns, # namespace in this module
          selectInput(
            ns("summary_fun_name"),
            "Select gene summary",
            names(summary_funs)
          )
        )
      ),
      output = plotOutput(ns("plot")),
      pre_output = pre_output,
      post_output = post_output
    )
  )
}
```

### Server function

Here we take the `summary_funs` too as we need to look up the function in it.

For the parsing of genes from text input we need another helper function first.

```{r}
extract_words <- function(x) {
  assert_string(x, min.chars = 1L)
  stringr::str_extract_all(
    x,
    "[a-zA-Z0-9:\\.]+"
  )[[1]]
}

# This seems pretty robust:
extract_words("a, b, , c, 234; 34562 - GeneID:bla")
extract_words("GeneID:1820, sdf.393; 32596")
```


```{r}
srv_g_signature <- function(input,
                            output,
                            session,
                            datasets,
                            mae_name,
                            summary_funs) {
  # When the filtered data set of the chosen experiment changes, update the
  # experiment data object.
  experiment_data <- reactive({
    req(input$experiment_name) # Important to avoid running into NULL here.

    mae <- datasets$get_data(mae_name, filtered = TRUE)
    mae[[input$experiment_name]]
  })

  # When the filtered data set or the chosen experiment changes, update
  # the call that creates the chosen experiment data object.
  experiment_call <- reactive({
    req(input$experiment_name) # Important to avoid running into NULL here.

    dat <- datasets$get_filtered_dataset(mae_name)
    dat$get_filter_states(input$experiment_name)$get_call()
  })

  # When the chosen experiment call changes, we recompute gene names.
  genes <- eventReactive(experiment_call(), ignoreNULL = FALSE, {
    object <- experiment_data()
    rownames(object)
  })

  # When the chosen experiment changes, recompute the assay names.
  assay_names <- eventReactive(input$experiment_name, ignoreNULL = TRUE, {
    object <- experiment_data()
    SummarizedExperiment::assayNames(object)
  })

  # When the assay names change, update the choices for assay.
  observeEvent(assay_names(), {
    assay_name_choices <- assay_names()

    updateSelectInput(
      session,
      "assay_name",
      choices = assay_name_choices,
      selected = assay_name_choices[1]
    )
  })

  # `reactiveValues` object for storing current gene text input.
  parsed_genes <- reactiveVal(NULL, label = "Parsed genes")

  # 3) the parsed genes are entered via text
  observeEvent(parsed_genes(), ignoreNULL = TRUE, {
    # Resolve reactives.
    genes <- genes()
    parsed_genes <- parsed_genes()

    # Only select the subset of parsed genes included in choices.
    new_selected <- intersect(parsed_genes, genes)
    removed <- setdiff(parsed_genes, new_selected)
    teal.widgets::updateOptionalSelectInput(
      session,
      "y_vars",
      choices = genes,
      selected = new_selected
    )
    n_removed <- length(removed)
    if (n_removed > 0) {
      showNotification(paste(
        "Removed", n_removed,
        ifelse(n_removed > 1, "genes", "gene"),
        parenthesize(shorten_list(removed))
      ))
    }
  })

  # When
  # 1) the genes are recomputed,
  # 2) the lock is switched off,
  # then update the choices and selections for genes in the UI.
  observeEvent(list(genes(), input$lock), {
    genes <- genes()
    lock <- input$lock
    old_selected <- input$y_vars

    if (!lock) {
      new_selected <- intersect(old_selected, genes)
      removed <- setdiff(old_selected, new_selected)
      teal.widgets::updateOptionalSelectInput(
        session,
        "y_vars",
        choices = genes,
        selected = new_selected
      )
      n_removed <- length(removed)
      if (n_removed > 0) {
        showNotification(paste(
          "Removed", n_removed,
          ifelse(n_removed > 1, "genes", "gene"),
          parenthesize(shorten_list(removed))
        ))
      }
    }
  })

  # Return the UI for a modal dialog with gene text input.
  dataModal <- function(example_list) { # nolint
    modalDialog(
      textInput(
        session$ns("gene_text"),
        "Enter list of genes",
        placeholder = example_list
      ),
      span(
        "Please enter a comma-separated list of gene IDs.",
        "(Note that genes not included in current choices will be removed)"
      ),
      footer = tagList(
        modalButton("Cancel"),
        actionButton(session$ns("ok"), "OK")
      )
    )
  }

  # Show modal when button is clicked.
  observeEvent(input$open_text_input, {
    gene_choices <- genes()
    example_list <- shorten_list(gene_choices)
    showModal(dataModal(example_list))
  })

  # When OK button is pressed, attempt to parse the genes from the text.
  # Remove the modal and display notification message.
  observeEvent(input$ok, {
    gene_text <- input$gene_text

    if (!nzchar(gene_text)) {
      showNotification(
        "Please enter at least one full gene ID.",
        type = "error"
      )
    } else {
      parse_result <- extract_words(gene_text)
      showNotification(paste(
        "Received", length(parse_result), "genes from text input",
        parenthesize(shorten_list(parse_result))
      ))
      parsed_genes(parse_result)
      removeModal()
    }
  })

  output$plot <- renderPlot({
    # Resolve all reactivity.
    experiment_data <- experiment_data()
    y_vars <- input$y_vars
    assay_name <- input$assay_name
    summary_fun_name <- input$summary_fun_name

    # Require which states need to be truthy.
    req(
      assay_name,
      # Note: The following statements are important to make sure the UI inputs have been updated.
      isTRUE(assay_name %in% SummarizedExperiment::assayNames(experiment_data)),
      cancelOutput = FALSE
    )

    # Validate and give useful messages to the user. Note: no need to duplicate
    # here req() from above.
    validate(need(
      hermes::is_hermes_data(experiment_data),
      "please use HermesData() on input experiments"
    ))
    validate(need(!is.null(y_vars), "please select at least one gene"))
    y_vars_not_included <- setdiff(y_vars, rownames(experiment_data))
    n_not_incl <- length(y_vars_not_included)
    validate(need(
      identical(n_not_incl, 0L),
      paste(
        n_not_incl,
        ifelse(n_not_incl > 1, "genes", "gene"),
        parenthesize(shorten_list(y_vars_not_included)),
        "not included, please unlock or change filters"
      )
    ))

    # Output.
    draw_signature(
      object = experiment_data,
      assay_name = assay_name,
      x_var = "COUNTRY",
      y_vars = y_vars,
      summary_fun = summary_funs[[summary_fun_name]],
      fun_name = summary_fun_name
    )
  })
}
```

### Sample module

```{r}
sample_tm_g_signature <- function() {
  mae <- hermes::multi_assay_experiment
  mae_data <- dataset("MAE", mae)
  data <- teal_data(mae_data)
  app <- init(
    data = data,
    modules = modules(
      static = {
        tm_g_signature(
          label = "signature",
          mae_name = "MAE"
        )
      }
    )
  )
  shinyApp(app$ui, app$server)
}
```

### Try out

```{r}
sample_tm_g_signature()
```

## For Production

### Additional summary functions

We can add similar functions to `hermes`.

```{r}
#' Generate gene signature using mean Z-score
#'
#' Function to calculate the mean Z-score from RNAseq assay matrix
#' (one-gene-per-row, one-sample-per-column)
#' @export
calc_meanZ <- function(x) { # nolint
  stopifnot(class(x) == "matrix")
  zmat <- apply(x, 1, function(y) if (sd(y) > 0) scale(y) else rep(NA, length(y)))
  zmean <- apply(zmat, 1, mean, na.rm = TRUE)
  names(zmean) <- colnames(x)
  return(zmean)
}

#' Generate gene signature using PC1
#'
#' Function to calculate the PC1 using principal component analysis from RNAseq
#' assay matrix (one-gene-per-row, one-sample-per-column)
#' @export
calc_PC1 <- function(x) { # nolint
  stopifnot(class(x) == "matrix")
  pc <- prcomp(t(x), center = TRUE)
  return(pc$x[, 1])
}
```




