---
title: "Barplot module design"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Install teal
remotes::install_github(
  "insightsengineering/teal",
  ref = "1185_dataset_specific_filter_panel",
  upgrade = "never",
  force = TRUE
)

library(teal)
library(teal.devel)
library(hermes)
library(tern)
```

## Barplot module design

We want a barplot teal module.

## Example data

```{r}
MAE <- multi_assay_experiment # from hermes # nolint
mae <- dataset("MAE", MAE)

data <- teal_data(mae)
```

## Plot function

In order to test the plotting separately, we write a separate function for the plot.
This takes the SE object, the name of the assay, the optional facet variable, the optional fill variable and the names of the x variables which are genes, and produces the ggplot.

```{r}
is.blank <- function(x) identical(x, "") # nolint

library(assertthat)
draw_barplot <- function(object, assayname, facetvar, fillvar, xvar, percentiles) {
  assert_that(
    is(object, "SummarizedExperiment"),
    is.string(assayname),
    is.null(facetvar) || is.string(facetvar),
    is.null(fillvar) || is.string(fillvar),
    is.string(xvar),
    is.vector(percentiles)
  )
  mat <- assay(object, assayname)

  # we add this since xvar, yvar could be blank first
  if (is.blank(xvar)) {
    xvar <- 1
  }
  df <- data.frame(x = mat[xvar, ])
  percentile_thresholds <- setdiff(percentiles, c(0, 1))
  df$final_x <- tern::cut_quantile_bins(df$x, probs = percentile_thresholds)
  cd <- colData(object)
  if (!is.null(facetvar)) {
    df$facet <- cd[[facetvar]]
  }
  if (!is.null(fillvar)) {
    df$fill <- cd[[fillvar]]
  }
  p <- ggplot(df, aes(x = final_x)) +
    geom_bar() +
    labs(x = xvar)
  if (!is.null(facetvar)) {
    p <- p + facet_wrap(~ facet)
  }
  if (!is.null(fillvar)) {
    p <- p + aes(fill = fill)
  }
  p
}
```

## Test it

```{r}
object <- HermesData(summarized_experiment)
```

```{r}
draw_barplot(
  object,
  assayname = "counts",
  facetvar = NULL,
  fillvar = NULL,
  xvar = genes(object)[1],
  percentiles = c(0.2, 0.8)
)
```

Here we need to pay attention to the fact that the quantiles are calculated once based on the entire dataset and are not recalculated once we add the faceting. This should be clearly reflected in the documentation to the end user. 

```{r}
object <- HermesData(summarized_experiment)
draw_barplot(
  object,
  assayname = "counts",
  fillvar = "SEX",
  facetvar = "AGE18",
  xvar = genes(object)[1],
  percentiles = c(0.4, 0.5)
)
```

## UI function

Next we define the user interface via the UI function.

Note that:
1. The experiment choices are taken directly from the (filtered) MAE, so we don't need to hardcode them.
2. The facetvar choices will be updated in the server function based on the chosen experiment.
3. Same for the assay.
4. Also the gene choices depend in the same way on the experiment, so follow the same flow. Since there are many choices here we use `selectizeInput()` see https://shiny.rstudio.com/articles/selectize.html for the genes.

```{r}
# helper function
emptySelectInput <- function(inputId, label, ...) { # nolint
  selectInput(
    inputId = inputId,
    label = label,
    choices = "",
    selected = "",
    ...
  )
}

ui_barplot <- function(id, datasets, dataname) {
  ns <- NS(id)
  mae <- datasets$get_data(dataname, filtered = TRUE)
  standard_layout(
    output = plotOutput(ns("plot")),
    encoding = div(
      tags$label("Encodings", class = "text-primary"),
      helpText("Analysis Data:", tags$code(dataname)),
      selectInput(ns("experiment"), "experiment", names(mae)),
      emptySelectInput(ns("assay"), "Select Assay"),
      selectizeInput(ns("xvar"), "Select Gene of Interest", choices = ""),
      sliderInput(ns("percentiles"),
                  "Select quantiles to be displayed",
                  min = 0, max = 1,
                  value = c(0.2, 0.8)
                  ),
      optionalSelectInput(ns("fillvar"), "Select stratify by"),
      optionalSelectInput(ns("facetvar"), "Select facet by")
    )
  )
}
```
## Server function

```{r}
srv_barplot <- function(input, output, session, datasets, dataname) {

  # Reactive for experiment data since that is used in multiple places below.
  experiment_data <- reactive({
    mae <- datasets$get_data(dataname, filtered = TRUE)
    experiment <- if (is.null(input$experiment)) {
      1
    } else {
      input$experiment
    }
    mae[[experiment]]
  })

  genes <- reactive({
    object <- experiment_data()
    rownames(object)
  })

  colvars <- reactive({
    object <- experiment_data()
    names(colData(object))
  })

  experiment_call <- reactive({
   dat <- datasets$get_filtered_dataset(dataname)
   if (is.null(input$experiment)) {
     NULL
   } else {
     dat$get_filter_queues(input$experiment)$get_call()
   }
  })

  # When the chosen experiment call changes,
  # we recompute gene names.
  genes <- eventReactive(experiment_call(), ignoreNULL = FALSE, {
   object <- experiment_data()
   rownames(object)
  })

  # When the chosen experiment changes, recompute the assay names.
  assaynames <- eventReactive(input$experiment, ignoreInit = TRUE, {
   object <- experiment_data()
   SummarizedExperiment::assayNames(object)
  })

  # When the chosen experiment changes, recompute the col vars.
  colvars <- eventReactive(input$experiment, {
   object <- experiment_data()
   names(colData(object))
  })

  # When the assay names change, update the choices for assay.
  observeEvent(assaynames(), {
    # First: resolve all reactivity.
    assaynames <- assaynames()

    # Second: do the action.
   updateSelectInput(
     session,
     "assay",
     choices = assaynames,
     selected = assaynames[1]
   )
  })

  # When the colvars change, update the choices for facetvar and fillvar.
  observeEvent(colvars(), {
    # First: resolve all reactivity.
    colvars <- colvars()

    # Second: do the action.
    idnames <- c("facetvar", "fillvar", "percentiles")
    for (i in seq_along(idnames)) {
      updateOptionalSelectInput(
        session,
        idnames[i],
        choices = colvars,
        selected = NULL
      )
    }
  })

  # When the genes are recomputed, update the choices for genes in the UI.
  observeEvent(genes(), {
    # First: resolve all reactivity.
    gene_choices <- genes()

    # Second: do the action.
    updateSelectizeInput(
      session,
      "xvar",
      choices = gene_choices,
      selected = gene_choices[1],
      server = TRUE
    )

  })

  output$plot <- renderPlot({
    # First: resolve all reactivity.
    experiment <- input$experiment
    object <- experiment_data()
    xvar <- input$xvar
    facetvar <- input$facetvar
    fillvar <- input$fillvar
    percentiles <- input$percentiles
    assayname <- ifelse(input$assay == "", "counts", input$assay)

    # Second: validate.
    validate(need(is.string(experiment), "please select one experiment"))
    validate(need(is.string(assayname), "please select one assay"))
    validate(need(is.string(xvar), "please select one gene for the x variable"))

    # Third: plot
    draw_barplot(
      object,
      assayname = assayname,
      xvar = xvar,
      facetvar = facetvar,
      fillvar = fillvar,
      percentiles = percentiles
    )
  })
}
```
Unfortunately in the current shiny version in the dockers we cannot yet try this out separately actually. With shiny above 1.5 it might work... we'll try this later.

## Teal module function

```{r}
tm_g_barplot <- function(label = "Simple MAE module",
                         info = NULL,
                         dataname = NULL,
                         pre_output = NULL,
                         post_output = NULL) {
  module(
    label = label,
    server = srv_barplot,
    ui = ui_barplot,
    ui_args = list(dataname = dataname),
    server_args = list(dataname = dataname),
    filters = "all"

  )
}
```

## Try it out

```{r}
app <- init(
  data = data,
  modules = root_modules(
    static = {
      tm_g_barplot(
        label = "barplot",
        dataname = "MAE"
      )
    }
  )
)

shinyApp(app$ui, app$server)
```
