---
title: "Gene signature module design"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(teal)
library(teal.modules.hermes)
library(hermes)
library(checkmate)
```

## Objectives

We need a shiny module that captures the gene signature logic. We might also need additional functions and classes which capture reused functionality. 

## Hermes Interface

### Helpers 

```{r}
shorten_list <- function(x) {
  n <- length(x)
  if (n > 3) {
    x <- c(x[1:2], "...", x[n])
  }
  paste(x, collapse = ", ")
}
shorten_list(letters[1:4])
shorten_list(letters)

parenthesize <- function(x) {
  if (is_blank(x)) {
    ""
  } else {
    paste0("(", x, ")")
  }
}

extract_words <- function(x) {
  assert_string(x, min.chars = 1L)
  stringr::str_extract_all(
    x,
    "[a-zA-Z0-9:\\.]+"
  )[[1]]
}

# This seems pretty robust:
extract_words("a, b, , c, 234; 34562 - GeneID:bla")
extract_words("GeneID:1820, sdf.393; 32596")
```

This will make our life easier below.

### Object idea

It seems that we should try to combine in a joint object, say named `GeneSpec`:
- `genes`: which genes are used.
- `summary_fun`: optional summary function which gives back one value per sample if `genes` contains more than one gene, or `NULL` which would mean don't summarize so use multiple genes (e.g. for the boxplot module we will need that)
- `summary_name`: string containing the name of the used summary function, defaults to `deparse(substitute(summary_fun))`

Probably this can be an R6 class.

```{r}
library(R6)

GeneSpec <- R6::R6Class( # nolint
  "GeneSpec",
  public = list(
    initialize = function(genes,
                          summary_fun = NULL,
                          summary_name = deparse(substitute(summary_fun))) {
      assert_character(genes, any.missing = FALSE, unique = TRUE, null.ok = TRUE)
      assert_function(summary_fun, null.ok = TRUE)
      assert_string(summary_name, min.chars = 1L)

      self$genes <- genes
      self$summary_fun <- summary_fun
      self$summary_name <- summary_name
    },
    get_genes = function() {
      self$genes
    },
    returns_vector = function() {
      identical(length(self$genes), 1L) || is.function(self$summary_fun)
    },
    get_label = function() {
      assert_true(self$returns_vector()) # only makes sense if vector valued
      if (length(self$genes) > 1) {
        paste0(self$summary_name, parenthesize(shorten_list(self$genes)))
      } else {
        self$genes
      }
    },
    extract = function(assay) {
      assert_names(rownames(assay), must.include = self$genes)
      assay_cols <- assay[self$genes, , drop = TRUE]
      if (length(self$genes) > 1 && is.function(self$summary_fun)) {
        summary_res <- self$summary_fun(assay_cols)
        assert_numeric(summary_res, len = ncol(assay_cols))
        summary_res
      } else {
        assay_cols
      }
    },
    validate = function(all_genes) {
      validate(need(!is.null(self$genes), "please select at least one gene"))
      genes_not_included <- setdiff(self$genes, all_genes)
      n_not_incl <- length(genes_not_included)
      validate(need(
        identical(n_not_incl, 0L),
        paste(
          n_not_incl,
          ifelse(n_not_incl > 1, "genes", "gene"),
          parenthesize(shorten_list(genes_not_included)),
          "not included, please unlock or change filters"
        )
      ))
    },
    genes = NULL,
    summary_fun = NULL,
    summary_name = NULL
  )
)
```

Let's try it out:

```{r}
x_spec <- GeneSpec$new(genes = "GeneID:1820")

x_spec$returns_vector()
x_spec$get_genes()
x_spec$get_label()

x_spec <- GeneSpec$new(genes = c("GeneID:1820", "GeneID:52"), summary_fun = colMeans)

x_spec$returns_vector()
x_spec$get_genes()
x_spec$get_label()
```


### Draw function

Just as an example here a simplified scatterplot function.

```{r}
draw_scatterplot2 <- function(object,
                              assay_name,
                              x_spec,
                              y_spec) {
  assert_class(object, "AnyHermesData")
  assert_string(assay_name)
  assert_class(x_spec, "GeneSpec")
  assert_class(y_spec, "GeneSpec")

  # make sure we get vectors back and not matrices with multiple genes
  assert_true(x_spec$returns_vector())
  assert_true(y_spec$returns_vector())

  # now use extract method on matrix which internally applies summary function
  # etc.
  assay_matrix <- assay(object, assay_name)
  df <- data.frame(
    x = x_spec$extract(assay_matrix),
    y = y_spec$extract(assay_matrix)
  )

  # then do the plot
  p <- ggplot(df, aes(x = .data$x, y = .data$y)) +
    geom_point() +
    labs(x = x_spec$get_label(), y = y_spec$get_label()) # we get the labels via class
  p
}
```

Let's try it out.

```{r}
hd <- HermesData(summarized_experiment)
draw_scatterplot2(
  hd,
  assay_name = "counts",
  x_spec = GeneSpec$new(genes(hd)[2]),
  y_spec = GeneSpec$new(genes(hd)[3:10], summary_fun = colMeans)
)

# nolint start
# draw_scatterplot2(
#   hd,
#   assay_name = "counts",
#   x_spec = gene_spec(genes(hd)[2]),
#   y_spec = gene_spec(genes(hd)[3:10], summary_fun = colMeans)
# )
# nolint end
```

So this works quite nicely and makes the `hermes` drawing functions light again.

## Shiny Module Prototype 1

So let's try to start our first shiny module for this.

### Module UI

In this case we just do input, no output, so we can name the module UI function accordingly.

```{r}
geneSpecInput <- function(id, # nolint
                          axis,
                          summary_funs) {
  ns <- NS(id)
  tagList(
    include_css_files(pattern = "*"),
    tags$div(
      class = "row",
      tags$div(
        class = "col-sm-8",
        tags$label(
          class = "control-label",
          paste("Select", axis, "gene(s)")
        )
      ),
      tags$div(
        class = "col-sm-4",
        actionButton(
          ns("open_text_input"),
          tags$span(icon("font fa-border")),
          title = "Enter list of genes",
          class = "list-genes"
        ),
        tags$div(
          title = "Lock gene selection",
          shinyWidgets::prettyToggle(
            ns("lock"),
            value = FALSE,
            label_on = NULL,
            label_off = NULL,
            status_on = "default",
            status_off = "default",
            outline = FALSE,
            plain = TRUE,
            icon_on = icon("lock fa-border"),
            icon_off = icon("unlock-alt fa-border"),
            animation = "pulse"
          )
        )
      )
    ),
    tags$div(
      class = "custom-select-input",
      teal.widgets::optionalSelectInput(
        ns("genes"),
        label = NULL,
        choices = "",
        multiple = TRUE
      )
    ),
    conditionalPanel(
      condition = "input.genes.length > 1", # JavaScript expression
      ns = ns, # namespace in this module
      selectInput(
        ns("summary_fun_name"),
        "Select gene summary",
        names(summary_funs)
      )
    )
  )
}
```

### Module Server

We can add some checkmate functions.

```{r}
assert_static <- function(x) {
  # to do : real implementation
  stopifnot(!is.reactive(x))
}

assert_reactive <- function(x) {
  # to do : real implementation
  stopifnot(is.reactive(x))
}
```

Here we take the `summary_funs` too as we need to look up the function in it.

```{r}
geneSpecServer <- function(id, # nolint
                           summary_funs,
                           genes) {
  assert_static(summary_funs)
  assert_list(summary_funs)
  assert_reactive(genes)

  moduleServer(id, function(input, output, session) {
    # `reactiveValues` object for storing current gene text input.
    parsed_genes <- reactiveVal(NULL, label = "Parsed genes")

    # the parsed genes are entered via text
    observeEvent(parsed_genes(), ignoreNULL = TRUE, {
      # Resolve reactives.
      genes <- genes()
      parsed_genes <- parsed_genes()

      # Only select the subset of parsed genes included in choices.
      new_selected <- intersect(parsed_genes, genes)
      removed <- setdiff(parsed_genes, new_selected)
      teal.widgets::updateOptionalSelectInput(
        session,
        inputId = "genes",
        choices = genes,
        selected = new_selected
      )
      n_removed <- length(removed)
      if (n_removed > 0) {
        showNotification(paste(
          "Removed", n_removed,
          ifelse(n_removed > 1, "genes", "gene"),
          parenthesize(shorten_list(removed))
        ))
      }
    })

    # When
    # 1) the genes are recomputed,
    # 2) the lock is switched off,
    # then update the choices and selections for genes in the UI.
    observeEvent(list(genes(), input$lock), {
      genes <- genes()
      lock <- input$lock
      old_selected <- input$genes

      if (!lock) {
        new_selected <- intersect(old_selected, genes)
        removed <- setdiff(old_selected, new_selected)
        teal.widgets::updateOptionalSelectInput(
          session,
          inputId = "genes",
          choices = genes,
          selected = new_selected
        )
        n_removed <- length(removed)
        if (n_removed > 0) {
          showNotification(paste(
            "Removed", n_removed,
            ifelse(n_removed > 1, "genes", "gene"),
            parenthesize(shorten_list(removed))
          ))
        }
      }
    })

    # Return the UI for a modal dialog with gene text input.
    dataModal <- function(example_list) { # nolint
      modalDialog(
        textInput(
          session$ns("gene_text"),
          "Enter list of genes",
          placeholder = example_list
        ),
        tags$span(
          "Please enter a comma-separated list of gene IDs.",
          "(Note that genes not included in current choices will be removed)"
        ),
        footer = tagList(
          modalButton("Cancel"),
          actionButton(session$ns("ok"), "OK")
        )
      )
    }

    # Show modal when button is clicked.
    observeEvent(input$open_text_input, {
      gene_choices <- genes()
      example_list <- shorten_list(gene_choices)
      showModal(dataModal(example_list))
    })

    # When OK button is pressed, attempt to parse the genes from the text.
    # Remove the modal and display notification message.
    observeEvent(input$ok, {
      gene_text <- input$gene_text

      if (!nzchar(gene_text)) {
        showNotification(
          "Please enter at least one full gene ID.",
          type = "error"
        )
      } else {
        parse_result <- extract_words(gene_text)
        showNotification(paste(
          "Received", length(parse_result), "genes from text input",
          parenthesize(shorten_list(parse_result))
        ))
        parsed_genes(parse_result)
        removeModal()
      }
    })

    # return the GeneSpec
    reactive({
      GeneSpec$new(
        input$genes,
        summary_fun = summary_funs[[input$summary_fun_name]],
        summary_name = input$summary_fun_name
      )
    })
  })
}
```


### Teal Module Usage

Here we allow the user to modify the list of gene signature functions.

```{r}
tm_g_signature <- function(label,
                           mae_name,
                           summary_funs = list(
                             Mean = colMeans,
                             Median = colMedians,
                             Max = colMaxs
                           ),
                           pre_output = NULL,
                           post_output = NULL) {
  assert_string(label)
  assert_string(mae_name)
  assert_list(
    summary_funs,
    types = "function",
    min.len = 1L,
    unique = TRUE,
    any.missing = FALSE,
    names = "unique"
  )
  assert_tag(pre_output, null.ok = TRUE)
  assert_tag(post_output, null.ok = TRUE)

  module(
    label = label,
    server = srv_g_signature,
    server_args = list(
      mae_name = mae_name,
      summary_funs = summary_funs
    ),
    ui = ui_g_signature,
    ui_args = list(
      mae_name = mae_name,
      summary_funs = summary_funs,
      pre_output = pre_output,
      post_output = post_output
    ),
    datanames = "all"
  )
}
```

#### UI function

Here we take the `summary_funs` list and give that into the option menu.

```{r}
ui_g_signature <- function(id,
                           datasets,
                           mae_name,
                           summary_funs,
                           pre_output,
                           post_output) {
  ns <- NS(id)
  mae <- datasets$get_data(mae_name, filtered = FALSE)
  experiment_name_choices <- names(mae)

  teal.widgets::standard_layout(
    encoding = tags$div(
      selectInput(ns("experiment_name"), "Select experiment", experiment_name_choices),
      selectInput(ns("assay_name"), "Select assay", choices = ""),
      geneSpecInput(ns("x_spec"), "x", summary_funs),
      geneSpecInput(ns("y_spec"), "y", summary_funs)
    ),
    output = plotOutput(ns("plot")),
    pre_output = pre_output,
    post_output = post_output
  )
}
```

#### Server function

```{r}
srv_g_signature <- function(input,
                            output,
                            session,
                            datasets,
                            mae_name,
                            summary_funs) {
  # When the filtered data set of the chosen experiment changes, update the
  # experiment data object.
  experiment_data <- reactive({
    req(input$experiment_name) # Important to avoid running into NULL here.

    mae <- datasets$get_data(mae_name, filtered = TRUE)
    mae[[input$experiment_name]]
  })

  # When the filtered data set or the chosen experiment changes, update
  # the call that creates the chosen experiment data object.
  experiment_call <- reactive({
    req(input$experiment_name) # Important to avoid running into NULL here.

    dat <- datasets$get_filtered_dataset(mae_name)
    dat$get_filter_states(input$experiment_name)$get_call()
  })

  # When the chosen experiment call changes, we recompute gene names.
  genes <- eventReactive(experiment_call(), ignoreNULL = FALSE, {
    object <- experiment_data()
    rownames(object)
  })

  # When the chosen experiment changes, recompute the assay names.
  assay_names <- eventReactive(input$experiment_name, ignoreNULL = TRUE, {
    object <- experiment_data()
    SummarizedExperiment::assayNames(object)
  })

  # When the assay names change, update the choices for assay.
  observeEvent(assay_names(), {
    assay_name_choices <- assay_names()

    updateSelectInput(
      session,
      "assay_name",
      choices = assay_name_choices,
      selected = assay_name_choices[1]
    )
  })

  x_spec <- geneSpecServer("x_spec", summary_funs, genes)
  y_spec <- geneSpecServer("y_spec", summary_funs, genes)

  output$plot <- renderPlot({
    # Resolve all reactivity.
    experiment_data <- experiment_data()
    assay_name <- input$assay_name
    x_spec <- x_spec()
    y_spec <- y_spec()

    # Require which states need to be truthy.
    req(
      assay_name,
      # Note: The following statements are important to make sure the UI inputs have been updated.
      isTRUE(assay_name %in% SummarizedExperiment::assayNames(experiment_data)),
      cancelOutput = FALSE
    )

    # Validate and give useful messages to the user. Note: no need to duplicate
    # here req() from above.
    validate(need(
      hermes::is_hermes_data(experiment_data),
      "please use HermesData() on input experiments"
    ))
    # Need to think about whether an object method is the best here or a separate
    # validation function that takes spec and genes.
    x_spec$validate(rownames(experiment_data))
    y_spec$validate(rownames(experiment_data))

    # Output.
    draw_scatterplot2(
      object = experiment_data,
      assay_name = assay_name,
      x_spec = x_spec,
      y_spec = y_spec
    )
  })
}
```

#### Sample module

```{r}
sample_tm_g_signature <- function() {
  mae <- hermes::multi_assay_experiment
  mae_data <- dataset("MAE", mae)
  data <- teal_data(mae_data)
  app <- init(
    data = data,
    modules = modules(
      tm_g_signature(
        label = "signature",
        mae_name = "MAE"
      )
    )
  )
  shinyApp(app$ui, app$server)
}
```

#### Try out

```{r}
sample_tm_g_signature()
```

## For Production

### hermes interface

we can add a constructor function e.g. `gene_spec` or `control_gene_spec` that avoids the
use of R6 syntax, see above commented code.

### Additional summary functions

```{r}
#' Generate gene signature using mean Z-score
#'
#' Function to calculate the mean Z-score from RNAseq assay matrix
#' (one-gene-per-row, one-sample-per-column)
#' @export
calc_meanZ <- function(x) { # nolint
  stopifnot(class(x) == "matrix")
  zmat <- apply(x, 1, function(y) if (sd(y) > 0) scale(y) else rep(NA, length(y)))
  zmean <- apply(zmat, 1, mean, na.rm = TRUE)
  names(zmean) <- colnames(x)
  return(zmean)
}

#' Generate gene signature using PC1
#'
#' Function to calculate the PC1 using principal component analysis from RNAseq
#' assay matrix (one-gene-per-row, one-sample-per-column)
#' @export
calc_PC1 <- function(x) { # nolint
  stopifnot(class(x) == "matrix")
  pc <- prcomp(t(x), center = TRUE)
  return(pc$x[, 1])
}
```

### Additional features for module

- make searchable for the symbol
- symbol first, then gene ID
- order alphabetically by symbol


