---
title: "Differential Expression module design"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(teal)
library(teal.devel)
library(hermes)
mae <- multi_assay_experiment # from hermes
data <- teal_data(dataset("MAE", mae))
```

## Scatterplot module design

We want to design a non-teal based differential expression module.

Work in design folder within an Rmd:
-plot function
-ui function
-server function
-teal module function
-try it out only interactively (not automatic / compiled)

Take as input the multi_assay_experiment in hermes and let the user select an experiment in the encoding
 -group by colData variable
 -thresholds selection for p-value and log FC
 -use diff_expression and autoplot from hermes to obtain volcano plot (no table yet at this point)
 -no reference/comparison group yet at this point, interactive plot, title settings, or Show R code for now.

Test the plot function for vocano plot:

```{r}
object <- HermesData(summarized_experiment)
colData(object) <- df_char_to_factor(colData(object))
res1 <- diff_expression(object, group = "SEX", method = "voom")
autoplot(res1, adj_p_val_thresh = 0.05, log2_fc_thresh = 2.5)
```

## UI function

```{r}
ui_volcanoplot <- function(id, datasets, dataname) {
  ns <- NS(id)
  mae <- datasets$get_data(dataname, filtered = TRUE)

  standard_layout(
    output = plotOutput(ns("plot")),
    encoding = div(
      selectInput(ns("experiment_name"), "Select experiment", names(mae)),
      selectInput(ns("facetvar"), "Facet by", choices = c("SEX" = "SEX", "AGE18" = "AGE18")),
      sliderInput(ns("log2_fc_thresh"), "Log2 folder change threshold", value = 2.5, min = 0.1, max = 10),
      sliderInput(ns("adj_p_val_thresh"), "Adjusted p-value threshold", value = 0.05, min = 0.01, max = 1)
    )
  )
}
```

## Server function

```{r}
srv_volcanoplot <- function(input, output, session, datasets, dataname) {

   experiment_data <- reactive({
     req(input$experiment_name)

     mae <- datasets$get_data(dataname, filtered = TRUE)

     object <- HermesData(mae[[input$experiment_name]])
     colData(object) <- df_char_to_factor(colData(object))
     object
   })

  diff_expr <- reactive({
    object <- experiment_data()
    facetvar <- input$facetvar

    req(facetvar)

    diff_expression(object, group = facetvar, method = "voom")
   })

  output$plot <- renderPlot({

    object <- experiment_data()
    diff_expr_result <- diff_expr()

    log2_fc_thresh <- input$log2_fc_thresh
    adj_p_val_thresh <- input$adj_p_val_thresh
    facetvar <- input$facetvar

    req(
      log2_fc_thresh,
      adj_p_val_thresh,
      facetvar
    )


    autoplot(diff_expr_result, adj_p_val_thresh = adj_p_val_thresh, log2_fc_thresh = log2_fc_thresh)
  })
}
```

## Teal module function

```{r}
tm_g_volcanoplot <- function(label = "Simple MAE module",
                             info = NULL,
                             dataname = NULL,
                             pre_output = NULL,
                             post_output = NULL) {
  module(
    label = label,
    server = srv_volcanoplot,
    ui = ui_volcanoplot,
    ui_args = list(dataname = dataname),
    server_args = list(dataname = dataname),
    filters = "all"
  )
}
```


## Try it out

Now we can try it out.

```{r}
app <- init(
  data = data,
  modules = root_modules(
    static = {
      tm_g_volcanoplot(
        label = "volcanoplot",
        dataname = "MAE"
      )
    }
  )
)
shinyApp(app$ui, app$server)
```

