---
title: "Scatterplot module design" 
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("https://raw.github.roche.com/gist/sabanesd/0e839ca7d4920fab342d8ed4b9d668fc/raw/eeaf752448e1b07641cc0c2a3af2172af7b99c94/install_nest.R")
install_nest("teal", "1185_dataset_specific_filter_panel")

library(teal)
library(teal.devel)
library(hermes)
```

## Scatterplot module design

We want to design a simple scatterplot teal module.

## Example data

```{r}
MAE <- multi_assay_experiment # from hermes
mae <- dataset("MAE", MAE)

data <- teal_data(mae) 
```


## Plot function

In order to test the plotting separately, we write a separate function for the plot.
This takes the SE object, the name of the assay, the optional facet variable, and the names of the x and y variables which are genes, and produces the ggplot.

Note: this is similar to https://github.roche.com/EDISNHL/dashdis/blob/master/R/draw_scatterplot.R
so we can look into that repo to get ideas.

```{r}
is.blank <- function(x) identical(x, "")

library(assertthat)
draw_scatterplot <- function(object, assayname, facetvar, xvar, yvar) {
  # browser()
  assert_that(
    is(object, "SummarizedExperiment"),
    is.string(assayname),
    is.null(facetvar) || is.string(facetvar),
    is.string(xvar),
    is.string(yvar)
  )
  mat <- assay(object, assayname)
  
  # we add this since xvar, yvar could be blank first
  if (is.blank(xvar)) {
    xvar <- 1
  }
  if (is.blank(yvar)) {
    yvar <- 1
  }
  df <- data.frame(x = mat[xvar, ], y = mat[yvar, ])
  if (!is.null(facetvar)) {
    cd <- colData(object)
    df$facet <- cd[[facetvar]]
  }
  p <- ggplot(df, aes(x = x, y = y)) +
    geom_point() +
    geom_smooth(method = "lm", formula = y ~ x) +
    labs(x = xvar, y = yvar)
  if (!is.null(facetvar)) {
    p <- p + facet_wrap(~ facet)
  }
  p
}
```

Test it:

```{r}
object <- HermesData(summarized_experiment)
draw_scatterplot(
  object,
  assayname = "counts",
  facetvar = "",
  xvar = genes(object)[1],
  yvar = genes(object)[2]
)
draw_scatterplot(
  object,
  assayname = "counts",
  facetvar = "SEX",
  xvar = genes(object)[1],
  yvar = genes(object)[2]
)
```

## UI function

Next we define the user interface via the UI function.

Note that:
1. The experiment choices are taken directly from the (filtered) MAE, so we don't need to hardcode them.
1. The facetvar choices will be updated in the server function based on the chosen experiment.
1. Same for the assay.
1. Also the gene choices depend in the same way on the experiment, so follow the same flow. 
1. Since there are many choices here we use `selectizeInput()` see https://shiny.rstudio.com/articles/selectize.html for the genes.

```{r}
ui_scatterplot <- function(id, datasets, dataname) {
  ns <- NS(id)
  mae <- datasets$get_data(dataname, filtered = TRUE)
  standard_layout(
    output = plotOutput(ns("plot")),
    encoding = div(
      tags$label("Encodings", class = "text-primary"),
      helpText("Analysis Data:", tags$code(dataname)),
      selectInput(ns("experiment"), "experiment", names(mae)),
      selectInput(ns("assay"), "assay", choices = ""),
      optionalSelectInput(ns("facetvar"), "facet by"),
      selectizeInput(ns("xvar"), "x gene", choices = ""),
      selectizeInput(ns("yvar"), "y gene", choices = "")
    )
  )
}
```

## Server function

```{r}
srv_scatterplot <- function(input, output, session, datasets, dataname) {
  
  # Reactive for experiment data since that is used in multiple places below.
  experiment_data <- reactive({
    mae <- datasets$get_data(dataname, filtered = TRUE)
    experiment <- if (is.null(input$experiment)) {
      1
    } else {
      input$experiment
    }
    mae[[experiment]]
  })
  
  experiment_call <- reactive({
    dat <- datasets$get_filtered_datasets(dataname)
    if (is.null(input$experiment)) {
      NULL
    } else {
      dat$get_filter_queues(input$experiment)$get_call()
    }
  })
  
  # When the chosen experiment call changes,
  # we recompute gene names.
  genes <- eventReactive(experiment_call(), ignoreNULL = FALSE, {
    object <- experiment_data()
    rownames(object)
  })
  
  # When the chosen experiment changes, recompute the assay names. 
  assaynames <- eventReactive(input$experiment, ignoreInit = TRUE, {
    object <- experiment_data()
    SummarizedExperiment::assayNames(object)
  })
  
  # When the chosen experiment changes, recompute the col vars.
  colvars <- eventReactive(input$experiment, {
    object <- experiment_data()
    names(colData(object))
  })
  
  # When the assay names change, update the choices for assay.
  observeEvent(assaynames(), {
    # First: resolve all reactivity.
    assaynames <- assaynames()

    # Second: do the action.
    updateSelectInput(
      session, 
      "assay",
      choices = assaynames,
      selected = assaynames[1]
    )
  })
  
  # When the colvars change, update the choices for facetvar.
  observeEvent(colvars(), {
    # First: resolve all reactivity.
    colvars <- colvars()

    # Second: do the action.
    updateOptionalSelectInput(
      session, 
      "facetvar",
      choices = colvars,
      selected = character()
    )
  })
  
  # When the genes are recomputed, update the choices for genes in the UI.
  observeEvent(genes(), {
    # First: resolve all reactivity.
    gene_choices <- genes()
    
    # Second: do the action.
    idnames <- c("xvar", "yvar")
    for (i in seq_along(idnames)) {
      updateSelectizeInput(
        session, 
        idnames[i],
        choices = gene_choices,
        selected = gene_choices[i],
        server = TRUE
      )
    }
  })
  
  output$plot <- renderPlot({
    # First: resolve all reactivity.
    experiment <- input$experiment
    object <- experiment_data()
    xvar <- input$xvar
    yvar <- input$yvar
    facetvar <- input$facetvar
    assayname <- input$assay
    
    # Second: validate.
    validate(need(is.string(experiment), "please select one experiment"))
    validate(need(is.string(assayname), "please select one assay"))
    validate(need(is.string(xvar), "please select one gene for the x variable"))
    validate(need(is.string(yvar), "please select one gene for the y variable"))
    validate(need(xvar != yvar, "please select different genes for x and y variables"))
    
    # Third: plot
    draw_scatterplot(
      object, 
      assayname = assayname, 
      xvar = xvar, 
      yvar = yvar, 
      facetvar = facetvar
    ) 
  })
}
```

Unfortunately in the current shiny version in the dockers we cannot yet try this out separately actually. With shiny above 1.5 it might work... we'll try this later.

## Teal module function

```{r}
tm_g_scatterplot <- function(label = "Simple MAE module", 
                             info = NULL, 
                             dataname = NULL, 
                             pre_output = NULL, 
                             post_output = NULL) {
  module(
    label = label,
    server = srv_scatterplot,
    ui = ui_scatterplot,
    ui_args = list(dataname = dataname),
    server_args = list(dataname = dataname),
    filters = "all"
  )
}
```

## Try it out

Now we can try it out.

```{r}
app <- init(
  data = data,
  modules = root_modules(
    static = {
      tm_g_scatterplot(
        label = "scatterplot",
        dataname = "MAE"
      )
    }
  )
)

shinyApp(app$ui, app$server)
```



