---
title: "Boxplot module design" 
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("https://raw.github.roche.com/gist/sabanesd/0e839ca7d4920fab342d8ed4b9d668fc/raw/eeaf752448e1b07641cc0c2a3af2172af7b99c94/install_nest.R")
install_nest("teal", "1185_dataset_specific_filter_panel")
library(teal)
library(teal.devel)
library(dplyr)
library(ggplot2)
library(random.cdisc.data)
library(hermes)
```

## Boxplot module design

We want to design a simple boxplot teal module.

## Example data

```{r}
MAE <- multi_assay_experiment # from hermes
mae <- dataset("MAE", MAE)

data <- teal_data(mae) 
```


## Plot function

In order to test the plotting separately, we write a separate function for the plot.
This takes the SE object, the name of the assay, the optional facet variable, and the names of the x variable which is sample information accessed using colData and y variable which is gene of interest, and produces the ggplot.


```{r}
is.blank <- function(x) identical(x, "")

library(assertthat)
draw_boxplot <- function(object, assayname, facetvar, xvar, yvar) {
  assert_that(
    is(object, "SummarizedExperiment"),
    is.string(assayname),
    is.null(facetvar) || is.string(facetvar),
    is.string(xvar),
    is.string(yvar)
  )
  
  mat <- assay(object, assayname)
  cd <- colData(object)
  # we add this since xvar, yvar could be blank first
  if (is.blank(xvar)) {
    xvar <- 1
  }
  if (is.blank(yvar)) {
    yvar <- 1
  }
  df <- data.frame(x = cd[, xvar], y = mat[yvar, ])
  if (!is.null(facetvar)) {
    df$facet <- cd[[facetvar]]
  }
  p <- ggplot(df, aes(x = x, y = y)) +
    geom_boxplot() +
    stat_boxplot(geom = "errorbar") +
    labs(x = xvar, y = yvar)
  if (!is.null(facetvar)) {
    p <- p + facet_wrap(~ facet)
  }
  p
}
```

Test it:

```{r}
object <- HermesData(summarized_experiment)
draw_boxplot(
  object,
  assayname = "counts",
  facetvar = NULL,
  xvar = "SEX",
  yvar = genes(object)[2]
)
draw_boxplot(
  object,
  assayname = "counts",
  facetvar = "RACE",
  xvar = "SEX",
  yvar = genes(object)[2]
)

```

## UI function

Next we define the user interface via the UI function.

Note that:
1. The experiment choices are taken directly from the (filtered) MAE, so we don't need to hardcode them.
2. The facetvar choices will be updated in the server function based on the chosen experiment.
3. Also the gene choices depend in the same way on the experiment, so follow the same flow. Since there are many choices here we should consider to use `selectizeInput()` see https://shiny.rstudio.com/articles/selectize.html in production.


```{r}

# helper function
emptySelectInput <- function(inputId, label, ...) {
  selectInput(
    inputId = inputId,
    label = label,
    choices = "",
    selected = "",
    ...
  )
}

ui_boxplot <- function(id, datasets, dataname) {
  ns <- NS(id)
  mae <- datasets$get_data(dataname, filtered = TRUE)
  standard_layout(
    output = plotOutput(ns("plot")),
    encoding = div(
      tags$label("Encodings", class = "text-primary"),
      helpText("Analysis Data:", tags$code(dataname)),
      selectInput(ns("experiment"), "experiment", names(mae)),
      selectInput(ns("assay"), "assay", choices = ""),
      selectizeInput(ns("yvar"), "Select Gene of Interest", choices = ""),
      selectizeInput(ns("xvar"), "Stratify by", choices = ""),
      optionalSelectInput(ns("facetvar"), "Facet by")
    )
  )
}
```

## Server function

```{r}
srv_boxplot <- function(input, output, session, datasets, dataname) {

  # Reactive function for experiment data since it is used in multiple places below.
  experiment_data <- reactive({
    mae <- datasets$get_data(dataname, filtered = TRUE)
    experiment <- if (is.null(input$experiment)) {
      1
    } else {
      input$experiment
    }
    mae[[experiment]] 
  })
  
  genes <- reactive({
    object <- experiment_data()
    rownames(object)
  })
  
  colvars <- reactive({
    object <- experiment_data()
    names(colData(object))
  })
  
  # When the colvars, update the choices for facetvar and stratification factor.
  # Todo: at the moment this always triggers when experiment data is updated.
  # how can we trigger this only when colvars really changes?
  observeEvent(colvars(), {
    # First: resolve all reactivity.
    colvars <- colvars()
    
    # Second: do the action.
    updateSelectInput(
      session,
      "xvar",
      choices = colvars,
      selected = ""
    )
    updateOptionalSelectInput(
      session, 
      "facetvar",
      choices = colvars,
      selected = ""
    )
  })
  
  # When the genes in the experiment change, update the choices for genes.
  # Todo: at the moment this always triggers when experiment data is updated.
  # how can we trigger this only when genes really changes?
  observeEvent(genes(), {
    # browser()
    # First: resolve all reactivity.
    genes <- genes()
    gene_choices <- if (is.null(genes)) {
      ""
    } else {
      genes
    }
    
    # Second: do the action.
      updateSelectInput(
        session, 
        "yvar",
        choices = gene_choices,
        selected = gene_choices[1]
      )
  })
    
  output$plot <- renderPlot({
    # browser()
    # First: resolve all reactivity.
    object <- experiment_data()
    xvar <- input$xvar
    yvar <- input$yvar
    facetvar <- input$facetvar
    assayname <- "counts"
    
    # Second: validate.
    # add later
    # teal.devel::as.global(experiment_data)
    
    # Third: plot
    draw_boxplot(
      object, 
      assayname = assayname, 
      xvar = xvar, 
      yvar = yvar, 
      facetvar = facetvar
    ) 
  })
}
```

Unfortunately in the current shiny version in the dockers we cannot yet try this out separately actually. With shiny above 1.5 it works:

## Teal module function

```{r}
tm_g_boxplot <- function(label = "Simple MAE module", 
                             info = NULL, 
                             dataname = NULL, 
                             pre_output = NULL, 
                             post_output = NULL) {
  module(
    label = label,
    server = srv_boxplot,
    ui = ui_boxplot,
    ui_args = args,,
    server_args = list(dataname = dataname),
    filters = "all"
  )
}
```

## Try it out

Now we can try it out.

```{r}
app <- init(
  data = data,
  modules = root_modules(
    static = {
      tm_g_boxplot(
        label = "boxplot",
        dataname = "MAE"
      )
    }
  )
)
shinyApp(app$ui, app$server)
```
